---
title: "Supplementary Material for "
author: "Maxime Dahirel"
date:
output: 
  html_document:
    theme: yeti
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE, warning=FALSE)
```

```{r packages-loading}
library(arm)

library(adegenet)

library(modelr)
library(RColorBrewer)

###inference
library(coda)
library(rstan)
library(brms)
rstan_options(auto_write = TRUE)
options(mc.cores = 2)


library(tidyverse)
library(here)
library(patchwork)

library(tidybayes)
library(bayesplot)

```

# S1 - Creation of genetic lines

*Trichogramma brassicae* wasps used in this experiment were derived from hybrid populations created in 2013 by mixing parental strains sampled in 24 sites different geographic sites in Western Europe.  These parental strains were initially collected for the purpose of a project on artificial selection and phenotype amelioration of *T. brassicae* strains for biological control. Isoline populations were then maintained under standard conditions between 2014 and 2016 (18 ± 0.5 °C, 65 ± 10% relative humidity, L:D 18:6), for approximately 20 generations. 

To generate independent experimental replicates with comparable levels of initial genetic diversity, we created three genetic mixes from three isoline populations each (using nine distinct isoline populations in total). We followed the protocol from Fellous *et al.*  [-@fellousCombiningExperimentalEvolution2014] to create mixes with homogeneous genetic background (i.e. similar representation of each of the three isolines). First, we performed all possible two-by-two crosses between the three isolines. The F1 progeny of each two-by-two cross was then crossed with the third remaining isoline. The F2 individuals from these crosses were then inbred for 2 generations to favour recombination. Individuals from these F4 were then mixed all together in equal proportions to form each final genetic mix. While they did differ from each other, all three mixes had broadly similar levels of genetic diversity at the start of the experiments (**Supplementary Figure S1.1**).

```{r supplementary-figure-1}
raw_genetics <- read_csv(here("data/Trichogramma_genetics.csv"), 
                         col_types = cols(.default=col_character())
)

data_genetics<-raw_genetics %>% 
  mutate( ### info on variables is contained in each sample ID
    Generation=str_extract(ID,"G0|G4|G8|G12"),
    Mix=str_extract(ID,"_1P|_2P|_3P"),
    Treatment=str_extract(ID,"PL|PS"),
    Replicate=str_extract(ID,"L_[:digit:]|S_[:digit:]")
  ) %>% 
  mutate(
    Generation=as.numeric(str_extract(Generation,"0|4|8|12")),
    Mix=as.numeric(str_extract(Mix,"1|2|3")),
    Replicate=as.numeric(str_extract(Replicate,"[:digit:]"))
  ) %>% 
  filter(Generation==0) %>% 
  mutate(Location="origin") %>% 
  mutate(landscapeID=paste("Mix",Mix,"_Treatment",Treatment,"_Replicate",Replicate,sep=""))

### STEP 2 isolate the genetics markers
genmat<- df2genind(select(data_genetics,starts_with("P")),####all allelic columns start with a "P"
            ncode=3,
            NA.char = "000",
            ind.names=data_genetics$ID,pop=data_genetics$landscapeID) #ncode=3 bc 3 characters for one allele

### STEP 3 generate summary of all other columns ; one row = one sampled patch
data_genetics <- data_genetics %>%   
  group_by(landscapeID) %>% 
  mutate(Nsampled = length(Location))%>%
  ungroup() %>% 
  select(-starts_with("P"), -ID) %>% 
  distinct() %>% 
  inner_join(
    tibble(Hexp=Hs(genmat),landscapeID=names(Hs(genmat)))
    )

data_genetics %>% 
  ggplot+
  geom_boxplot(aes(x=factor(Mix),y=Hexp),alpha=0.2)+
  geom_jitter(aes(x=Mix,y=Hexp),pch=20,size=4)+
  scale_y_continuous(name = "Genetic diversity at the start of the experiment (exp. heterozygosity)",limits=c(0.1,0.45)) + 
  scale_x_discrete("Genetic Mix")+
  theme_bw()
```

**Supplementary Figure S1.1** Genetic diversity (multilocus expected heterozygosity) as a function of genetic mix at the start of the experiment. One dot = one experimental landscape. The y axis is scaled to show the full range of genetic diversities observed during the course of the experiment, for reference (see main text)

# S2 - Validating computer estimates of *Trichogramma* population sizes

For analyses asking for population sizes (Supplementary Material 3 and 8), the proportion of parasitized eggs was estimated based on egg strip photographs using ImageJ [@abramoffImageProcessingImageJ2004] and the `CODICOUNT` plugin [@perezMethodeAnalyseImage2017].  `CODICOUNT` estimates the number of pixels assigned to parasitized host eggs, unparasitized host eggs and background after training. We tested each photograph using 4 macros, each trained on a different set of pictures (from generations 1 2, 3 and 13 respectively) and each having its own systematic bias, combined to better approximate real rates. Proportions of pixels identified as “egg” that were also identified as “parasitized” were used as our estimate of population size / parasitism rate.

To assess the validity of these automated counts, we manually counted egg plates from 197 patches (out of 2725 photographed for the experiment) roughly spanning the full range of possible population sizes (human count range: 17 to 443; number of hosts per population ≈ 450). 

```{r data-validate}

raw_dynamics<- read_csv(here("data","Trichogramma_dynamics.csv"))

data_validate<- raw_dynamics %>% 
  mutate(Mix= as.numeric(str_sub(Image,1,1)),
         Dynamics = str_sub(Image,2,3),
         Replicate = as.numeric(str_sub(Image,4,4)),
         Patch = as.numeric(str_sub(Image,6,7))
  ) %>% 
  #no further processing needed; we exploit the fact that chars 6&7 are either digit dot (patchs 0 to 9)
#or digit-digit (patches 10 and beyond)
## as.numeric() resolves both correctly (e.g; "2." and "12" become 2 and 12)
  mutate(landscapeID=paste("Mix",Mix,"_Dynamics",Dynamics,"_Replicate",Replicate,sep="")) %>% 
  mutate(Peggs_est=P/(P+H), ## proportion of pixels counted as parasitised (estimated)
         spacetimeID=paste(landscapeID,"_Generation",Generation,"_Patch",Patch,sep=""), # a unique ID for each replicate x patch x generation combination
         Mix = factor(Mix)) %>% 
  group_by(spacetimeID) %>% 
  filter(sum(is.na(obs_count)==FALSE)>0) %>% ### we group by spacetimeID (replicate, patch, time) and we filter out all cases without a human count to validate against
  mutate(Neggs_obs=replace_na(obs_count,0)) ### there are 4 macro counts but only one human count per spacetimeID (put on the same row as one of the macro counts); the other three are NAs, need to contain filler values (0 here) for the bivariate model to work, which won't be used in the analyses)
```

```{r model-validate}
prior_validate=c(
    set_prior("normal(0,1.5)",class="Intercept",resp="Neggsobs"),
    set_prior("normal(0,1)",class="sd",resp="Neggsobs"),
    
    set_prior("normal(0,1.5)",class="Intercept",resp="Peggsest"),
    set_prior("normal(0,1)",class="sd",resp="Peggsest"),  ##half-normal is implied for sd pars in brms
    set_prior("normal(0,1)",nlpar="invphi",resp="Peggsest",lb=0),  ##but has to be made explicit for many other types of pars
    set_prior("lkj(2)",class="cor")
)

bf_validate_obs=bf(Neggs_obs|trials(450)+subset(Neggs_obs>0)~(1|1|spacetimeID),family=binomial)
bf_validate_estimated=bf(Peggs_est~(1|1|spacetimeID),
                         nlf(phi~1/invphi),
                         invphi~1,
                         family=Beta(link_phi="identity"))

mod_validate=brm(
    mvbf(bf_validate_obs+bf_validate_estimated),
    data=data_validate,
    chains=4,iter=4000,prior=prior_validate,seed=42
)

cor_latent <-posterior_samples(mod_validate) %>% select(.value=contains("cor")) %>% mean_hdi()

newdata <- filter(data_validate,Macro=="MacroG3") %>% ##selecting a Macro at random, it's just for generating predictions and plots
  ungroup() %>% 
  mutate(Neggs_obs=2,prop_para_pix=0.2,Nhosts=450) %>% 
  select(Nhosts,Peggs_est,Neggs_obs,spacetimeID,Macro) %>% 
  distinct()

preds <- left_join(
  newdata %>% add_fitted_draws(mod_validate,resp="Peggsest",value="estimated",re_formula=~(1|spacetimeID)),
  newdata %>% add_fitted_draws(mod_validate,resp="Neggsobs",value="human",re_formula=~(1|spacetimeID))
) %>% 
  mutate(human=human/450)

cor_obs<-preds %>% 
  group_by(.draw) %>% 
  nest() %>% 
  mutate(.value= map(data,~cor.test(.$estimated,.$human)$est)) %>% 
  unnest(.value) %>%
  select(.draw,.value) %>% 
  ungroup() %>% 
  mean_hdi()

diff_obs_est<-preds %>% 
  mutate(diff=human-estimated) %>% 
  group_by(.draw) %>% 
  summarise(diff=100*mean(diff)) %>% 
  mean_hdi()
```

We then used a bivariate generalized linear mixed model approach to compare computer estimates and human counts, with a Beta family used for the former and a binomial distribution for the latter. This approach accounts for the data structure and the fact both human- and computer-derived counts are likely estimated with error [@perezMethodeAnalyseImage2017]. It allows us to estimate the correlation between human counts and “consensus” computer counts using information from all macros. Both models included a random intercept of population/patch identity, which were correlated.  Estimated and human-observed parasitism rates were highly correlated both on the latent logit scale (*r*= `r round(cor_latent$.value,2)` [`r round(cor_latent$.lower,2)`; `r round(cor_latent$.upper,2)`]) and on the observed scale (*r* = `r round(cor_obs$.value,2)` [`r round(cor_obs$.lower,2)`; `r round(cor_obs$.upper,2)`]). There was also no evidence of bias (predicted difference between average observed and estimated parasitism rates, in % of hosts per patch = `r round(diff_obs_est$diff,2)`% [`r round(diff_obs_est$.lower,2)`; `r round(diff_obs_est$.upper,2)`])

We then refitted this model, this time adding a fixed effect of the macro in the "computer estimates" part of the model, to determine whether macros presented systematic biases. We found that the overall absence of bias of the "consensus" estimates (see above) actually resulted from most macros slightly underpredicting "true" parasitism rates while one consistently overpredicted them (**Supplementary Figure S2.1**).

```{r model-validate2}
prior_validate=c(
    set_prior("normal(0,1.5)",class="Intercept",resp="Neggsobs"),
    set_prior("normal(0,1)",class="sd",resp="Neggsobs"),
    
    set_prior("normal(0,1.5)",class="b",resp="Peggsest"),
    set_prior("normal(0,1)",class="sd",resp="Peggsest"),  ##half-normal is implied for sd pars in brms
    set_prior("normal(0,1)",nlpar="invphi",resp="Peggsest",lb=0),  ##but has to be made explicit for many other types of pars
    set_prior("lkj(2)",class="cor")
)

bf_validate_obs=bf(Neggs_obs|trials(450)+subset(Neggs_obs>0)~(1|1|spacetimeID),family=binomial)
bf_validate_estimated=bf(Peggs_est~0+Macro+(1|1|spacetimeID),
                         nlf(phi~1/invphi),
                         invphi~1,
                         family=Beta(link_phi="identity"))

mod_validate2=brm(
    mvbf(bf_validate_obs+bf_validate_estimated),
    data=data_validate,
    chains=4,iter=10000,prior=prior_validate,seed=42
)


obs<- data_validate %>% ungroup() %>% 
  mutate(spacetimeID=spacetimeID[1],Neggs_obs=1) %>% 
  select(spacetimeID,Macro,Neggs_obs) %>% distinct() %>%  
  add_fitted_draws(mod_validate2,resp="Neggsobs",re_formula = NA)  %>% 
  mean_hdi()

data_validate %>% ungroup() %>% 
  mutate(spacetimeID=spacetimeID[1],Neggs_obs=1) %>% 
  select(spacetimeID,Macro,Neggs_obs) %>% distinct() %>%  
  add_fitted_draws(mod_validate2,resp="Peggsest",re_formula = NA) %>% 
  ggplot()+
  geom_hline(aes(yintercept=obs$.value[1]/450),size=1.2)+
  geom_hline(aes(yintercept=obs$.lower[1]/450),lty=2)+
  geom_hline(aes(yintercept=obs$.upper[1]/450),lty=2)+
  stat_eye(aes(Macro,.value))+
  scale_y_continuous("Parasitism rate", limits = c(0.25,0.75))+
  theme_bw()
```

**Supplementary Figure S2.1** Posterior mean estimated parasitism rate, by macro, for the 197 plates used to compare human and computer counts. Horizontal lines denote the mean (bold line) and 95% credible interval (dotted lines) for the posterior mean parasitism rate based on human counts.

# S3 – Validation of the experimental design

To determine whether our experimental reduction of connectedness did limit individual movement (i.e. functional connectivity), we used data from the first generation of testing, i.e. the only time the patch of origin of all parents can be known with certainty. We used parasitism rates to calculate for each landscape × macro the average egg location (in patches from release) and used it as a proxy of dispersal. We analysed this distance using a Beta generalised linear mixed model (dividing it by 4, the maximal possible distance, to ensure it stayed between 0 and 1). The model included a fixed effect of treatment, as well as a random effect of macro (intercept) to account for uncertainty in our estimates of parasitism and distances, more specifically for macro-level systematic biases. We additionally controlled for the total number of eggs laid in a landscape. Indeed, we expect eggs are more likely to have been laid farther in more fecund replicates, simply because the room in nearer patches is limited, rather than because of differences in connectedness. We summed patch-level parasitism rates for each landscape to obtain a measure of the total number of eggs, and included it (centred and scaled to unit SD) as a fixed effect in the model described above.
We found that released wasps did lay their eggs closer from release sites in landscapes with restricted connectedness compared to controls (**Supplementary Figure S3.1**). In addition, we confirmed higher fecundity led to higher egg-laying distances (see model summary in attached code).

```{r data-G0}
raw_dynamics<- read_csv(here("data","Trichogramma_dynamics.csv"))

data_popsize_temp<- raw_dynamics %>% 
  mutate(Mix= as.numeric(str_sub(Image,1,1)),
         Treatment = str_sub(Image,2,3),
         Replicate = as.numeric(str_sub(Image,4,4)),
         Patch = as.numeric(str_sub(Image,6,7))
  ) %>% 
  #no further processing needed; we exploit the fact that chars 6&7 are either digit dot (patchs 0 to 9)
#or digit-digit (patches 10 and beyond)
## as.numeric() resolves both correctly (e.g; "2." and "12" become 2 and 12)
  mutate(landscapeID=paste("Mix",Mix,"_Treatment",Treatment,"_Replicate",Replicate,sep="")) %>% #a unique replicate ID
  mutate(Peggs_est=P/(P+H), ## proportion of pixels counted as parasitised (estimated)
         spacetimeID=paste(landscapeID,"_Generation",Generation,"_Patch",Patch,sep=""), # a unique ID for each replicate x patch x generation combination
         Mix = factor(Mix)) %>% 
  mutate(Treatment = fct_recode(Treatment,`control`="PL",`restricted connectedness`="PS"))

data_G0 <- filter(data_popsize_temp,Generation==1 & Patch<=4) %>%  
  ##only patches 0 to 4 are available at start (adding "Patch <=4" above should not be necessary, but it makes it explicit)
  select(landscapeID,Mix,Treatment,Patch,Peggs_est,Macro) %>% 
  ungroup() %>% 
  pivot_wider(.,names_from="Patch",values_from=c("Peggs_est"),names_prefix = "P") %>% 
  filter(is.na(P0)==FALSE) %>%   ## remove one row with the one macro with missing info for "resident" eggs (can't estimate dispersal)
  mutate(P1=replace_na(P1,0)) %>% 
  mutate(P2=replace_na(P2,0)) %>% 
  mutate(P3=replace_na(P3,0)) %>% 
  mutate(P4=replace_na(P4,0))  %>% 
  mutate(Pall=(P0+P1+P2+P3+P4)) %>% 
  mutate(distmean = (1*P1+2*P2+3*P3+4*P4) / Pall, ## mean distance from release
         scale_Pall = scale(Pall)) %>%   
  mutate(jitteredTRT=jitter(as.numeric(factor(Treatment)))) %>% ## just for pretty plotting
  group_by(landscapeID) %>% mutate(jitteredTRT=mean(jitteredTRT)) %>%  ##same: ensures same jitter for same measure from different macros
  ungroup()
```


```{r model-g0}

mod_G0 <- brm(bf(distmean/4~0+Treatment+scale_Pall+(Treatment+scale_Pall|Macro), 
                 ###divided by 4 because it's the maximal possible distance possible at G0
        nlf(phi~1/invphi),
        invphi~1),
        data=data_G0,family=Beta(link_phi="identity"),
        iter=2000,chains=4,
        prior=c(
          set_prior("normal(0,1.5)",class="b"),
          set_prior("normal(0,1)",class="sd"),
          set_prior("normal(0,1)",nlpar="invphi",lb=0)
        ),
        control=list(adapt_delta=0.99),seed=42)
```

```{r figure-G0}

newdata <- data_G0 %>% 
  mutate(Macro=Macro[1],scale_Pall=0) %>%  #it's just a placeholder, we plot the posterior for the "true" (grand) mean distance, without random effects
  select(Macro, Treatment,scale_Pall) %>% 
  distinct()

## these averages are just indicative and for plotting purposes
data_G0_avg <- data_G0 %>% 
  group_by(landscapeID,Treatment)%>%  
  summarise(distmean=mean(distmean),
            jitteredTRT=mean(jitteredTRT)) %>% ungroup()

## the main plot                                                           
p1 <-newdata %>% 
  add_fitted_draws(mod_G0,re_formula=NA) %>% 
  ggplot()+
  stat_eye(aes(x=Treatment,y=.value*4,fill=Treatment),.width=c(0.66,0.95),slab_alpha=0.8,point_interval = mean_hdi) +
  ## we put both observed points (by macros) and the indicative averages for each each landscape
  geom_point(data=data_G0,aes(x=jitteredTRT,y=distmean,col=Treatment),fill="white",pch=21,alpha=0.8)+
  geom_point(data=data_G0_avg, aes(x=jitteredTRT,y=distmean,fill=Treatment),pch=21,size=2)+
  #scale_fill_manual(values=TRTpalette[c(1,4)])+
  #scale_colour_manual(values=TRTpalette[c(1,4)])+
  scale_x_discrete("")+
  scale_y_continuous("Mean egg-laying distance from release site (patches)",limits=c(-0.2,2.5))+
  theme_bw()

## a subplot showing the difference between treatments. Keep it or put this info in text?
p2<- newdata %>% 
  add_fitted_draws(mod_G0,re_formula=NA) %>% 
  ungroup() %>% 
  compare_levels(variable=.value,by=Treatment) %>% 
  ggplot()+
  stat_eye(aes(x=Treatment,y=.value*4),.width=c(0.66,0.95),fill="grey",point_interval = mean_hdi)+
  scale_x_discrete("",labels="")+
  scale_y_continuous("Difference between treatments",
                     limits=c(-0.2,2.5)-4*invlogit(fixef(mod_G0)["Treatmentcontrol","Estimate"])) + ##so the "difference" subplot is aligned with the main
  geom_hline(yintercept=0, lty=2) +
  theme_bw()

p1 + p2 + plot_layout(guides="collect",widths=c(4,1)) & theme_bw() &
  theme(legend.position="none")


##aside: very roughly showing that there is indeed consistent macro-level bias even on distances
## data_G0 %>% group_by(landscapeID) %>% mutate(DD=mean(distmean)) %>% ggplot()+ geom_boxplot(aes(x=Macro,y=distmean/DD))
## likely reason: because macro that over/underestimates parasitism rates will over/underestimate # of eggs in "other patches" so over/underestimate distances
  
```

**Supplementary Figure S3.1.** Mean egg-laying distance from release patch as a function of treatment. Shaded areas: posterior distribution (assuming total fecundity fixed to its average); black dots and segment: posterior mean and 66/95% credible intervals; empty colored dots: estimated values for each macro; full colored dots: averages across macros for each replicate.


# S4 - Detailed description of models (main text)

We here describe the full structure of the models presented in the main text, as well as associated prior choices. We follow throughout the same notation conventions as McElreath [-@mcelreathStatisticalRethinkingBayesian2020].Variable names are reset from one model to the next, for simplicity.

## S4.1 - Genetic diversity, experimental data

We can describe the expected heterozygosity $H_{x,i,t}$ of a patch in replicate $i$, with $x$ denoting the location of that patch (0 for core patches, 1 for front patches) and $t$ the number of generations since release, the following way:

$$ 
\begin{equation}
\tag{S4.1-1}
logit(H_{x,i,t}) \sim StudentT(logit(\mu_{x,i,t}) , \sigma_{d} , \nu ) 
\end{equation}
$$
where $\mu_{x,i,t}$ is the mean genetic diversity, $\sigma_{d}$ the residual standard deviation and $\nu$ the number of degrees of freedom. 

Based on theory, we expect average genetic diversity to decline exponentially with time: 

$$ 
\begin{equation}
\tag{S4.1-2}
\mu_{x,i,t} =   H_{i,t=0} \times  \exp(-\lambda_{x,i} \times t) 
\end{equation}
$$

The two parameters $H_{i,t=0}$ (initial diversity) and  $\lambda_{x,i}$ (rate of decline) depend on replicate, treatment (control or restricted connectedness) and location (core or front) as follow:

$$
\begin{equation}
\tag{S4.1-3}
logit(H_{i,t=0}) = \beta_{0} + \alpha_{i} 
\end{equation}
$$

$$
\begin{equation}
\tag{S4.1-4}
log_{10}(\lambda_{x,i}) = \beta_{1[TREATMENT]}+\beta_{2[TREATMENT]}(x-0.5) + \gamma_{i} + \zeta_{i}(x-0.5)
\end{equation}
$$
(with $logit$ and $log_{10}$ transformations here to ensure both parameters stay within bounds).
Initial diversity is thus independent of treatment and location, and only depends on replicate identity through a random effect $\alpha_{i}$, while the rate of decline $\lambda$ depends on treatment, location (here centred, which makes it possible to fit the model even though data at t=0 are neither from the front or the core) and their interaction. In addition, both the average decline rate and the front-core difference depend on replicate, which is accounted for through the random effect parameters $\gamma_{i}$ and $\zeta_{i}$ respectively.

The random effects are distributed as following: 

$$
\begin{equation}
\tag{S4.1-5}
\begin{bmatrix}
\alpha_{i} \\
\gamma_{i} \\
\zeta_{i} \\
\end{bmatrix} \sim 
MVNormal(
\begin{bmatrix}
0\\
0\\
0\\
\end{bmatrix},\boldsymbol{\Omega})
\end{equation}
$$
with $\boldsymbol{\Omega}$ the covariance matrix:

$$
\begin{equation}
\tag{S4.1-6}
\boldsymbol{\Omega} = 
\begin{bmatrix}
\sigma_{\alpha} & 0 & 0 \\
0 & \sigma_{\gamma} & 0 \\
0 & 0 & \sigma_{\zeta} \\
\end{bmatrix} 
\times \boldsymbol{R} \times
\begin{bmatrix}
\sigma_{\alpha} & 0 & 0 \\
0 & \sigma_{\gamma} & 0 \\
0 & 0 & \sigma_{\zeta} \\
\end{bmatrix}
\end{equation}
$$
where $\sigma_{\alpha}$, $\sigma_{\gamma}$ and $\sigma_{\zeta}$ are the random effect standard deviations for each parameter, and $\boldsymbol{R}$ the corresponding correlation matrix.

We used weakly informative priors mostly following McElreath [-@mcelreathStatisticalRethinkingBayesian2020]:

$$ 
\begin{equation}
\tag{S4.1-7}
\beta_{0} \sim Normal(0,1.5) 
\end{equation}
$$
as it corresponds to the logit of a proportion. For all other fixed effect parameters:

$$
\begin{equation}
\tag{S4.1-8}
\beta \sim Normal(0, 1) 
\end{equation}
$$
(note that results are overall insensitive to fixing both standard deviations to 1 or to 1.5).


All standard deviation parameters (including the distributional standard deviation $\sigma_{d}$) were attributed the same prior:

$$ 
\begin{equation}
\tag{S4.1-9}
\sigma \sim halfNormal(0,1) 
\end{equation}
$$
the random effect correlation matrix a prior mildly skeptical of high correlations:

$$
\begin{equation}
\tag{S4.1-10}
R \sim LKJCorr(2)
\end{equation}
$$
while we used 

$$
\begin{equation}
\tag{S4.1-12}
\nu \sim Gamma(2,0.1)
\end{equation}
$$

for the degrees of freedom based on Stan developers recommendations (https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations).

An alternative model can be written, where $H_{x,i,t} \sim Beta(\mu_{x,i,t}, \phi)$, using the (mean, precision) parametrisation of the Beta distribution, but it has a lower predictive performance (see Methods in the main text, and associated supplementary data and code). Following recommendations for weakly informative priors (https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations), we use a half-Normal prior on the inverse of the precision parameter $\phi$ :

$$
\begin{equation}
\tag{S4.1-13}
1/\phi\sim halfNormal(0,1)
\end{equation}
$$


### S4.2 - Genetic diversity, simulated data

In our simulations, it is not possible to analyse genetic diversity using the model(s) described above, in large part because simulated diversity data contain many zeroes, which mean models using either logit transformations or Beta distributions are inappropriate.

As an alternative, we cas use the model proposed by Gandhi *et al.* [-@gandhiCooperationMitigatesDiversity2019a], in which the expected among-replicate variance in the frequencies of either one of the two alleles $V_{j,t}$, for a given treatment and location combination $j$ at a given time $t$ (in generations from release), can be described as following:

$$ V_{j,t} \sim Beta(\mu_{j,t}, \phi)$$

$$ \mu_{j,t} =   V_{max} \times  (1 - \exp(-\lambda_{j} \times t)) $$ 

where $V_{max}$ is both the maximum possible variance and the initial genetic diversity (and such, only depends on the initial allelic distribution, not on treatment), and $\lambda$ a rate of decay of genetic diversity, which varies between treatments and locations (core vs. front) as in the "experimental" model.

We put the same priors on $log_{10}(\lambda)$ and $1/\phi$  as in the experimental models ($Normal(0, 1)$ and $halfNormal(0, 1)$, respectively) , for the same reasons. We put a stronger, informative prior on $logit(V_{max})$ :

$$ logit(V_{max}) \sim Normal(logit(0.25), 1.5) $$
, as the expected asymptotic among-replicate variance (and the initial heterozygosity) is 0.25 for the case of a randomly distributed and randomly fixating locus with two alleles (the variance of a Bernoulli-distributed variable with *p*= 0.5).

## S4.3 - Front velocity, experimental and simulated data

In both experimental and simulation data, the location $X_{i,t}$ of the expansion front of the $i$th replicate after $t$ generations can be modelled as follow:

$$ X_{i,t} \sim logNormal(v_{i,t} \times t, \sigma_{d}) $$

where $v_{i,y}$ denotes the expansion speed since release, and $\sigma{d}$ is the residual standard deviation of the log-distances.

The expansion speed $v_{i,y}$ can be modelled as following: 

$$ v_{i,t} = v_{i,t=\infty} + (v_{i,t=1} - v_{i,t=\infty}) \times \exp(-\lambda_{treatment,i} \times (t -1)) $$
  , that is, moving from an initial speed $v_{i,t=1}$ to an equilibrium speed $v_{i,t=\infty}$, with the rate of "decay" to the asymptic speed being $\lambda_{treatment,i}$.
  
All three parameters depend on experimental/simulation treatment, and on replicate identity, in the same way:

$$
\ln(v_{i,t=1}) = ln(v_{TREATMENT[t=1]}) + \alpha_{i}
$$

$$
\ln(v_{i,t=\infty}) = ln(v_{TREATMENT[t=\infty]}) + \gamma_{i}
$$

$$ log_{10}(\lambda_{i}) = log_{10}(\lambda_{TREATMENT}) + \zeta_{i} $$

The random effects $\alpha_{i}$, $\gamma_{i}$ and $\zeta_{i}$ are distributed as following: 

$$
\begin{bmatrix}
\alpha_{i} \\
\gamma_{i} \\
\zeta_{i} \\
\end{bmatrix} \sim 
MVNormal(
\begin{bmatrix}
0\\
0\\
0\\
\end{bmatrix},\boldsymbol{\Omega})
$$

with $\boldsymbol{\Omega}$ the covariance matrix:

$$
\boldsymbol{\Omega} = 
\begin{bmatrix}
\sigma_{\alpha} & 0 & 0 \\
0 & \sigma_{\gamma} & 0 \\
0 & 0 & \sigma_{\zeta} \\
\end{bmatrix} 
\times \boldsymbol{R} \times
\begin{bmatrix}
\sigma_{\alpha} & 0 & 0 \\
0 & \sigma_{\gamma} & 0 \\
0 & 0 & \sigma_{\zeta} \\
\end{bmatrix}
$$

where $\sigma_{\alpha}$, $\sigma_{\gamma}$ and $\sigma_{\zeta}$ are the random effect standard deviations for each parameter, and $\boldsymbol{R}$ the corresponding correlation matrix.

We used the same weakly informative priors $Normal(0,1)$ for the fixed effects parameters $v_{TREATMENT[t=1]}$, $v_{TREATMENT[t=\infty]}$ and $\lambda_{TREATMENT}$, $halfNormal(0,1)$ priors for all standard deviations $\sigma$ and again a LKJ prior $LKJcorr(2)$ for the correlation matrix $\boldsymbol{R}$


# S5 - main text model coefficient tables
TO DO

# S6 - propagule size at front in simulations
To understand how range expansions in simulated landscapes with reduced connectedness could be pushed (based on velocity ratios) and yet lose diversity faster than controls, we looked at the population sizes of newly established front patches, immediately after immigration and before reproduction. We focused on “control” and “restricted connectedness” treatments for simplicity, and compared the mean population size at establishment of a new front patch (corresponding to the size of the “propagule” sent from the previous front patch) as a function of treatment and equilibrium population size K, as well as their interactions. To do so, we used a negative binomial GLMM with the aforementioned variables as fixed effects and replicate identity as a random effect.
We found that the average propagule size was lower in reduced connectedness landscapes (Supplementary Figure S6), likely explaining why genetic diversity decayed faster, as the effective population size would also be lower. However, we also noted that while average dispersal rate was reduced by half in “reduced connectedness” landscapes, propagule size was reduced by less than that (Supplementary Figure S6). This, to our opinion, solves the apparent paradox of restricted connectedness: these waves do send more individuals to the front than expected based on their base dispersal rate, explaining why they moved faster than their Fisher velocity and hence are pushed. However, at the dispersal rates we considered, this is not enough to compensate the negative effect of the lower dispersal rate itself, hence the faster decay of genetic diversity.

Supplementary Figure S6.  Posterior distribution of the mean size Nnew of newly founded populations in simulations, estimated using a negative binomial mixed model.

# S7 - new treatment in simulations DDD + reduced connectedness
To tentatively explain why, in our experiments, pushed wave fronts created by reduced connectedness lost less genetic diversity than controls, when our first simulations indicate they should not, we ran an additional simulation scenario. In this scenario, both the base dispersal rate and the slope of the density-dispersal relationship were changed compared to the control (D0 = 0.1, βdensity = 1), leading to a reaction norm where dispersal rates are reduced by 50% at low densities, but similar to the controls at population sizes close to K (DK ≈ 0.23). We ran this scenario under the same conditions of K and number of replicates as the scenarios presented in the main text, and re-fitted the statistical models presented in the main text with it.
We show that this new scenario leads to pushed waves, based on relative velocities (Supplementary Figure X). We also show that, contrary to the simple “reduced connectedness” scenario, it results in a slower loss genetic diversity than in controls (Supplementary Figure XX). Interestingly, compared to the simple “reduced connectedness” scenario again, this one resulted in absolute front velocities much closer to controls (Supplementary Figure X), a feature that better matches results from experimental landscapes.
FIGURE LEGENDS TO ADD (Figures next page)

# S8 - Differences in equilibrium population size between experimental treatments

We used a Beta generalized linear mixed model to determine whether *Trichogramma* population sizes (estimated parasitism rates) differed between our two experimental treatments. We focused on core patches that were expected to be at equilibrium; for the purposes of this analysis, we conservatively defined “core” patches as “patches for which the distance to the release site is equal or lower than 25% of the distance release site – front patch at the time of sampling”. Our model included fixed effects of experimental treatment, as well as random effects of observation macro (to account for systematic macro bias, see **Supplementary Figure S2.1**), replicate identity, and patch identity nested in replicate to account for repeated measures within a given experimental replicate and across time.

We found that equilibrium population sizes in the range core were higher when connectedness was restricted than in controls (121.7% on average [106.25; 137.3] of the average population size of controls, Supplementary Figure S8).

Supplementary Figure S8 Posterior mean population sizes (host parasitism rates, each population harbours ≈ 450 host eggs) as a function of landscape type.

# S9 - Model descriptions for models presented in Supplementary Materials

## S9.1 - validation of codicount (Supplementary Material S2)

TO DO

## S9.2 -  validation of reduced connectedness (Supplementary Material S3)

$$ D_{[treatment,macro]} \sim logNormal(\mu, \sigma) $$

$$\mu = \mu + \alpha $$

or

$$\mu = \mu + \beta \times fecundity + \alpha $$

## S9.3 - propagule size model (Supplementary Material S6)

TO DO

## S9.4 - new simulation treatment (Supplementary Material S7)

See S4.2 and S4.3

## S9.5 core population size model (Supplementary Material S8)

$$ P_{[treatment,macro,x,t]} \sim Beta(\mu_{[treatment,x,t]}, \phi) $$

$$ \mu_{[treatment,x,t]} = \mu_{[treatment]} +\alpha_{[i]}  +\gamma_{[i]}$$


# References
