---
title: 'Trichogramma range expansions (genetics and dynamics): 2- individual-based model, model creation'
author: "Maxime Dahirel"
date:
output: 
  html_document:
    theme: yeti
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

```{r load-packages}
library(nlrx)
library(arm)
library(tidyverse)
library(cowplot)
library(here)
library(patchwork)
```

# Introduction

(see manuscript/preprint text for more context)

Pushed expansion waves are generated when spread is positively density-dependent, i.e. when dispersal and/or growth is positively density-dependent. Pushed waves have mostly been studied using the latter (which corresponds to Allee effects), but recent works have started to focus on density-dependent dispersal, as it is common in nature. We propose to expand on the effects of variable dispersal by looking at stochasticity in dispersal. We predict that (at low population sizes) reduced connectivity should lead to increased stochasticity in dispersal success and establishment difficulties, creating positive density-dependent spread through another way.

The aim of the model is to test this prediction using information on front velocity and genetic diversity in the front and core patches of expanding metapopulations. To validate the simulation, we also set up treatments with Allee effects and (classical) density-dependent dispersal. We focus on relatively "low" population sizes, as they are where our effect of interest is present, and have been understudied in theory. We look at two populations sizes because of previous research showing equilibrium population size *K* influences the dynamics of pushed waves, and to have a qualitative point of comparison in case experimental data also involve shifts in *K*.

# Setup

First, let's tell R where is your NetLogo version, what version number it is, for how many generations do we want the model to run, and how many replicates we want for each parameter combination:

```{r general-information}
####
NLPATH <- "C:/Program Files/NetLogo 6.1.1"
NLVERSION <- "6.1.1"
nreplicates <- 100
duration <- 100
```

We now need to tell R where the actual NetLogo file of the model is (The reader is strongly encouraged to go look at it, especially the Info pane):

```{r model-path}
nl_object <- nl(
  nlversion = NLVERSION, nlpath = NLPATH,
  modelpath = here("NetLogo_model/range_expansion.nlogo"),
  jvmmem = 1024
)
```


Then, let's set the experimental design. There are four main treatments, plus a combination ( density-dependent dispersal + reduced connectivity) that is used in Supplementary Materials. More detailed information on parameter and parameter choices can be found in the manuscript and in the Info pane of the NetLogo file

```{r design-grid}
### designing the experiment
trt_grid <- tibble(
  treatment = c(
    "reference", "weak Allee effect (a = 0.95)",
    "density-dependent dispersal",
    "reduced connectivity",
    "reduced + DDD"
  ),
  start_allee_thres = c(0, 0.95, 0, 0, 0), ## weak Allee effect
  slope_disp_mean = c(0, 0, 1, 0, 1), ## leads to roughly doubling of dispersal between 0 and K, realistic
  disp0_mean = c(0.2, 0.2, 0.2, 0.1, 0.1) ## realistic range
) %>%
  mutate(fecundity = 5) %>% ## more or less arbitrary, but inspired by insect/Trichogramma biology, see manuscript
  expand_grid(K = c(225, 450)) %>%
  mutate(
    v_f = 2 * sqrt( ## v_f = expected velocity of "equivalent" pulled wave:
      log(fecundity) * (1 - 1 / K) * (1 - start_allee_thres / 1) * ## fecundity term at N = 1
        0.5 * invlogit(logit(disp0_mean) + (1 / K) * slope_disp_mean)
    ) ## dispersal term at N = 1
  )
```

The corresponding growth and dispersal functions are plotted Figure 2:

```{r figure2}
trt_grid2 <- trt_grid %>% 
  expand_grid(N=c(1:450)) %>% 
  mutate(dispersal_rate=invlogit(logit(disp0_mean)+(N/K)*slope_disp_mean),
         mean_fecundity=exp(log(fecundity)*(1- (N/K) )*(1- (start_allee_thres/N)))) %>% 
  mutate(K=factor(K)) %>% 
  mutate(treatment = relevel(factor(treatment), ref = "reference")) %>%
  mutate(treatment = fct_relevel(treatment, "reduced connectivity", after = Inf)) %>%
  mutate(treatment = fct_relevel(treatment, "reduced + DDD", after = Inf))

plot_disp<-ggplot(trt_grid2)+
  geom_line(aes(x=N,y=dispersal_rate,group=K,lty=K)) +
  geom_point(data=trt_grid2 %>% filter(N==1),aes(x=N,y=dispersal_rate,group=K)) +
  scale_y_continuous("mean dispersal rate",limits = c(0,0.75)) +
  scale_x_continuous("") +
  facet_grid(cols=vars(treatment))  +
  theme_half_open(11) +
  background_grid(colour.major = "grey95", colour.minor = "grey95")

plot_fec<-ggplot(trt_grid2)+
  geom_line(aes(x=N,y=mean_fecundity,group=K,lty=K)) +
  geom_point(data=trt_grid2 %>% filter(N==1),aes(x=N,y=mean_fecundity,group=K)) +
  scale_y_continuous("mean per capita fecundity") +
  scale_x_continuous("Population size N") +
  facet_grid(cols=vars(treatment))  +
  theme_half_open(11) +
  background_grid(colour.major = "grey95", colour.minor = "grey95")

plot_disp/plot_fec + plot_layout(guides = 'collect') + plot_annotation(tag_levels="A")
```



Now we need to add all of this (plus information on the variables that stay constant, which output variables will we save...) to the experiment slot of the NL object (we encourage readers to look at the info pane in the .nlogo model file for details):

```{r set-experiment}

nl_object@experiment <- experiment(
  expname = "experiment-art1-2020",
  outpath = here("NetLogo_output"),
  repetition = 1,
  tickmetrics = "true",
  idsetup = "setup",
  idgo = "go",
  stopcond = "not any? turtles",
  runtime = duration,
  evalticks = 1:duration,
  metrics = c("ticks"),
  metrics.patches = c("pxcor", "N_predispersal", "N_postdispersal", "N_allele0", "N_allele1"),
  constants = list(
    "trait_variation" = "\"reshuffled\"", ### note how string variable/csts must be entered ### traits are redrawn at random every generation
    ## so no evolution . Moot because no trait variation (see below), but still
    # "K"=450,
    # "start_allee_thres"=5,
    "duration" = duration,
    "logit_disp0_sd" = 0, # individual variation in underlying trait, fixed to 0 for this paper
    # "disp0_mean"=0.2,
    "slope_disp_sd" = 0 # individual variation in underlying trait, fixed to 0 for this paper
    # "slope_disp_mean"=0,
    # "fecundity"=5,
  ),

  variables = list( ### needs to be a nested list
    "start_allee_thres" = list(values = trt_grid$start_allee_thres),
    "slope_disp_mean" = list(values = trt_grid$slope_disp_mean),
    "disp0_mean" = list(values = trt_grid$disp0_mean),
    "fecundity" = list(values = trt_grid$fecundity),
    "K" = list(values = trt_grid$K)
  )
)
```


And we're go for running the simulations! We set a "distinct" simulation design (see `nlrx` help).
(Careful, script is not parallelized, so running 500 replicates in total is expected to take up to a few hours)

```{r running-simulations}

set.seed(1989) ## we set a seed here to guarantee the seeds selected below are the same everytime
nl_object@simdesign <- simdesign_distinct(nl = nl_object, nseeds = nreplicates)
#nl_object@simdesign@simseeds<-1:nreplicates #or we can rewrite the nseeds here (to guarantee they are all different)
length(unique(nl_object@simdesign@simseeds))==nreplicates #check that all generated seeds are different
exp1_2020 <- run_nl_all(nl_object)

# setsim(nl_object, "simoutput") <- exp1_2020 #not needed
# write_simoutput(nl_object)   #not needed
```

# Output

We can now process the result for the main analyses (and their 5-treatment equivalent in supplementary material):

```{r processing-output-presaving}

tab <- exp1_2020 %>%
  dplyr::select(ticks, start_allee_thres, slope_disp_mean, disp0_mean, fecundity, K, seedID = `random-seed`, metrics.patches) %>%
  mutate(metrics.patches2 = map(metrics.patches,
    .f = ~ .x %>%
      filter(N_postdispersal > 0) %>% ## we only keep populated patches...
      filter(pxcor == max(pxcor) | pxcor == 0) %>% ## we only keep the edge and the starting patch
      mutate(is.edge = (pxcor == max(pxcor))) %>%
      mutate(
        Hexp = 2 * (N_allele0 / N_postdispersal) * (N_allele1 / N_postdispersal), # expected heterozygosity
        P0 = (N_allele0 / N_postdispersal),
        P1 = (N_allele1 / N_postdispersal)
      )
  )) %>%
  left_join(trt_grid) %>%
  select(-metrics.patches) %>%
  unnest(cols = c(metrics.patches2)) %>%
  mutate(replicateID = paste(K, treatment, seedID)) %>%
  mutate(Location_full = paste(replicateID, pxcor, ticks))
```


```{r saving-output}
write_csv(tab, path = here("NetLogo_output/model-output.csv"))
```

We can now use the `ibm-analysis` scripts to analyse the model outputs.
